Demystifying the Computer Science Ph.D. Admission in the US

### 2023-06-06 phd-cs-us.tex
Updates from Overleaf

\newcommand{\tool}{\textsf{NeuralSAT}}
\def\Section{\S}

\newcommand{\mycomment}[3][\color{black}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{Demystifying the Computer Science Ph.D. Admission in the US}
\date{}
\author{ThanhVu (Vu) Nguyen with inputs from faculty at GMU and other universities}
%\setcounter{secnumdepth}{3}
\begin{document}
\maketitle

\begin{abstract}
Having been involved in PhD admissions for many years and after
numerious interaction international (especially Vietnamese) students, I've
realized that international students lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted is difficult compared to CS PhD programs in other countries.

So I want to share my opinions and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While I wrote this document for CS applicants, it should be applicable to various STEM fields.
Moreover, while many examples given are for GMU, which is ranked 33rd on CSRankings (June 2016), the writing should be generalize to most other R1 universities  (though \emph{very} top schools migh be very extreme, e.g., see the admission process at \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{CMU}.

@@

If you have suggestions or comments, feel free to create a \href{https://github.com/dynaroars/dynaroars.github.io/issues}{GitHub issue} for discussion.
\end{abstract}

@@

\section{How is your application evaluated?}

After you submit your PhD application (usually in December), it will be first screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your writters submit their letters?
Then your application will be reviewed by a
\textbf{PhD admission committee} consisting of faculty members in CS. Each application is reviewed by about \emph{three} faculty members, who will evaluate your profile and try to reach a consensus about your case.

Many PhD committee members are \emph{new assistant professors}. This helps junior faculty, who typically don't have students yet, the opportunities to recruit students. The chair of the committee will be a senior professor, but they will not review individual applicaitions and instead assign them to committee members. The chair will look at various factors such as keywords/interests, research areas, or mentioning faculty names, and assign the application to appropriate faculty.

@@

In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. If the application process were free for everyone, we would receive an overwhelming number of applications to review.

@@

The primary focus of the admissions committee is to \textbf{evaluate your background and interest in research} since a PhD in Computer Science
is a research degree. To assess your research capability, we consider
the following key indicators.

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: Vietnamese students often mention Scopus Q1, which consists of diversed journals from IEEE, Elsevier, and several other publishers unfamiliar to me.  Honestly, I don't know/recognize the majority of journals listed in Scopus Q1. So this might be something to be mindful of, as CS faculty in the US might not be too familiar with Scopus or journals listed in there, so devote sometime in your statement to discuss their significance.
\end{tcolorbox}

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Craig: GMU and many other universities allow you to upload your published papers. In many cases, even if the papers were not published at top places, we can still determine their quality by simply skimming over the paper.  
\end{tcolorbox}

@@

Finally, this is something very easy to do, but is missed by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply there ? name several professors who you're
interested in working with (in some cases your application will be sent to them for evaluation).
Be careful not to send wrong statement to wrong school or mixing
facts (e.g., talking about school X but mentioned about working with
profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
should.

@@

Graduating from top universities \emph{that we recognize} helps (but still not as much as having research experience).
However, committee members might not know much about schools in your country and just treat your school as
\emph{``unknown foreign''}, which can be a minus point (if your school is well-known, then it is \emph{``top foreign''}, which is definitely a plus).

So what can you do about this? several things such as ask your CS dept to put itself on CSRankings (it's the easiest way to get CS people to know about the school), explain about your school in your statement (and ask your LOR writer to do that), and of course, if you're Vietnamese, consider a CS PhD program that has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese professors}.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: Sometime PhD admission committee will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. I have looked at Vietnamese applications (whether they are assigned to me or not) and provide inputs to the reviews of those applications, e.g., X is the top tech school in Vietnam and so it should be /top/ instead of /unknown/ foreign, which makes a huge difference.
\end{tcolorbox}

@@

This can be an issue for students in many international top universities 
where the competition is so high that very good students students can still have low ranking (and be overlooked by Admission committee).
So what to do with this? well, same as \hyperref[sec:your-school]{before}, e.g., put a note about this in your statement.

@@

You also do need to
take standarized English test. Just do well enough to pass minimum
requirement set by the university.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: The minimum for GMU CS (being above this might not mean much, but below is a \red{red flag}).
\begin{itemize}
\item GPA: $\ge 3.0$ in your undergrad
  \begin{itemize}
  \item but also consider the rank/prestige of your schools
  \end{itemize}
\item GRE: not required, most CS graduate programs no longer require GRE
%    \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
\item English requirement tests (one of the below)
  \begin{itemize}    
  \item TOEF: 88 pts in total AND $\ge 20$ points in each subsection OR
  \item IELTS: $\ge 6.5$ OR
  \item DuoLingo Graduate English: $\ge 120$ OR 
  \item Pearson Test of Academic English: $\ge 67$
  \end{itemize}  
\end{itemize}
\end{tcolorbox}

@@

An interview lasts about 20--30 mins, and one implicit thing you will be evaluated on is whether you can communicate effectively (i.e., speak/understand English).  You will also get chance to ask questions about the university so think of something to ask (just the same as you interview at a company).

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
At GMU, we are encouraged to interview candidates. In most cases we interview very strong candidates, i.e., the interview is actually to recruit them.  In some rare cases a faculty interviews a candidate that they see potentials and want argue for admission, i.e., without the interview, that candidate is definitely rejected. In any case, getting interview means you have a very good chance of getting admitted.
\end{tcolorbox}

@@

Most schools will have an \textbf{Open House}, which is a great resource to learn about the dept, faculty, area, etc. During the Open House, you get a chance to talk directly to multiple faculty and current students.  Take notes of faculty who make you excited, count those that are taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisors, the dept, the area, funding situation etc.  Ask about anything you want to determine that they deserve \emph{you}.

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
GMU has Virtual Open House, e.g., \url{https://cs-gmu.github.io/cs-phd-voh-s23/}, which I've co-organized in the last two years. We invite all admitted PhD students to the VOH to learn about the CS program, the department, GMU, and the DC area in general. Students also get opportunies to chat with professors and current students.  Everything is through Zoom.
\end{tcolorbox}

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD (in any fields) are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is the type of funding paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come the school, a company, or organization. Tab.~\ref{tab:funding} summarizes the differences.

\begin{table}
  \centering
  \caption{Different types of fundings}\label{tab:funding}
  \begin{tabular}{c|c|c|c}
    \toprule
    &\textbf{TA}&\textbf{RA}&\textbf{Fellowship}\\
    \midrule
    \textbf{From} & School & Profs. & School/External\\
    \textbf{For}                  & Teaching Assistant       & Research                        & Research                              \\
    \textbf{Tuition/Ins./Stipend} & Yes                      & Yes                             & Yes                                   \\
    \textbf{Cover Summer?}              & No                       & Maybe                           & Yes                                   \\
    \midrule
    \textbf{Pros}                 & Research Freedom         & Get to do research              & Research Freedom                      \\
    \textbf{Cons}                 & TA, Uncertain            & Research restriction, Uncertain & Competitive, limited             \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Graduating Assistantship (TA/RA)}
The most common type of funding is \textbf{graduate assistanship}, which is either TA or RA. Both TA and RA come with tuition waiving (you don't have to pay tuition), health insurance (this takes care of your insurance, which is a must have in the US), and most importantly, your stipend (i.e., your salary). Some universities also pay insurance for spouse/chilren (or give very good discount)

A couple of notes about stipend. First, the amount of stipend depends on the university, which in turns depend on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, a school year is (typically) 9-month in the US, so stipend is for 9 months (so divide by 9 for each month). Third, like for most source of income in the US, you will have to pay tax on your stipend. Finally, private universities might pay more for stipend.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
At GMU, TA and RA have similiar benefits in tuition waiving and insurance.  For stipend, depending on the college and department, a 9-month graduate assistant stipend is set.  TA and RA will usually be that ammount (TA will definitely be that, RA might flunctuate a bit depending on the stage of the student and the prof.). 
\end{tcolorbox}

@@

TA is common in the beginning when you haven't found your advisor who would pay you RA. As a TA, you help professors with their classes (e.g., grading or teaching labs/recitation). Your TAship is paid through the department, i.e., they hire you to help teach.  During a semester, a TA mgith work with several courses and professors (not necessary their advisor).  TA funding typically is not available during the summer, which has no school.

\paragraph{How to get TA?}  Unless you have other fundings such as RA or Fellowships, TA is typically a default thing. When you apply to be a full-time student, you also sate that you need financial assistant. It is common that the PhD committee will either admit you and give you GTA, or reject you; i.e., we do not admit a student without supporting them.

@@

Even if you have other fundings and do not need TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for research career where you often have to give talks and tell people about your work. Note that GMU sometimes has classes that a more seninor student can teach.  In that case, you will be paid as a lecturer, which is higher than GTA.  This is a good opportunity for students to get teaching experience and also get paid more.
\subsection{Research Assistant (RA)}
RA is provided through a professor through their own funding so you can work on their project.  
You do not need to teach as an RA, so you can focus on your research. Depending on the professor, RA may be available during the summer.

\textbf{How to get RA?} If a professor recruits you, they will likely give you RA right away (e.g., when you apply).  A common scenario is that you get admitted with TA, and then after a year or two find an advisor to support you with RA.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: If you're lucky and got recruited by a prof. who would give you RA right away, it's very likely you will get admitted.  For example, if a prof., even if not in PhD admissin committee, wants to work with and funds you, the PhD admission committee will respect that decision and admit your application (unless your application has many red flags).
\end{tcolorbox}

@@

Fellowship is another type of funding in which the student applies for (e.g., from school, industries, government). Fellowships are typically competitive and generous, and gives pretty much all benefits tuition/insurance that a graduate assistantship has.  Moreover, it often gives higher stipend (including summer) and opens doors for job opportunities (e.g., internship).  For example, a student with a Microsoft fellowship will likely get an internship at Microsoft.

@@

\textbf{How to get Fellowship?} You apply for them.  The US government has many fellowships but these likely require US citizenship or residency.  However, tech companies including Google, Microsoft, Facebook, IBM all have fellowships that international students can apply for. Prestigious fellowships typically require a clear and good research plan, so it is usueful to have research experience and even publication for a better chance (remember, you're competiting with the top Ph.D. students at top universities worldwide).

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
At GMU, Ph.D. applicants are automatically eligible for a Presidential Fellowship.  It is at least as good as GTA but the most important thing is that as a fellowship it is trully free money (i.e., you are not depending on any prof. or TA duties).  PhD admission committee members nominate applicants for this fellowship and the committee will vote and give the fellowship to the top 2.
\end{tcolorbox}

@@

\begin{itemize}
    \item   Is an MS degree required for admission to PhD?
    \begin{description}[before=\small]
    \item No. In fact, student with BS can get MS degree "along the way" to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.
    \end{description}

    \item Can I apply to PhD to CS if my undergrad was in Math, Biology, Finance, etc?
    \begin{description}[before=\small]
        \item Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned.
    \end{description}

    \item What can you do to increase your admission chance?
    \begin{description}[before=\small]
        \item Show something that makes you \textbf{stand out}, e.g., are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS?  All of these are important in CS in the US.

@@

    \item How do I address a professor Firstname Lastname?
    \begin{description}[before=\small]
        \item If you don't know that professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste the names, so no need to do so,  just Prof. or Dr. Lastname.

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
    Vu: I've been called Dr. Vu and I find it a bit amusing but am totally fine with it.
    \end{tcolorbox}
    \end{description}
\end{itemize}

% **

>>>>

\def\Section{\S}

\newcommand{\mycomment}[3][\color{black}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\title{Demystifying the Computer Science Ph.D. Admission in the US}
\date{}
\author{ThanhVu (Vu) Nguyen, George Mason University}
%\setcounter{secnumdepth}{3}
\begin{document}
\maketitle

\begin{abstract}
Having been involved in PhD admissions for many years and after
numerous interaction with  students (especially Vietnamese), I've
realized that \emph{international} students lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted is difficult compared to CS PhD programs in other countries.

So I want to share my opinions and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While I wrote this document for CS applicants, it should be applicable to various STEM fields.
Moreover, while many examples given are for GMU, which is 33rd on CSRankings (June 2016), the writing should be generalize to most other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US refers to a research-intensive university with a high level of research activity across various disciplines.} universities  (though \emph{very} top schools might be very extreme, e.g., see the admission process at \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{CMU}).

@@

If you have suggestions or comments, feel free to create a \href{https://github.com/nguyenthanhvuh/phd-cs-us/issues}{GitHub issue} for discussion.
\end{abstract}

@@

\section{How is Your Application Evaluated?}

After you submit your PhD application (usually in December), it will be first screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters?
Then your application will be reviewed by a
\textbf{PhD admission committee} consisting of faculty members in CS. Each application is assigned to about \emph{three} faculty members, who will evaluate your profile and try to reach a consensus about your case.  Note that while in most cases the assigned reviewers will be the main ones deciding your application, every faculty will have access to your application and can give inputs on your profile.

In many cases, the admission committee will involve junior faculty (e.g., assistant professors) in the department. While this requires  a significant time investment, it provides junior faculty the opportunities to recruit students. The chair of the committee will be a senior professor, but they will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names, and assign the applications to appropriate faculty.

@@

In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Also, if the application process were free for everyone, we would receive an overwhelming number of applications to review.

@@

The primary focus of the admissions committee is to \textbf{evaluate your background and interest in research} since a PhD in Computer Science
is a research degree. To assess your research capability, we consider
the following key indicators, listed in order of importance.

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: Vietnamese students often mention Scopus Q1, which consists of diverse journals from IEEE, Elsevier, and several other publishers unfamiliar to me.  Honestly, I don't know/recognize the majority of journals listed in Scopus Q1. So this might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed in there, so devote sometime in your statement to discuss the significance of your papers.
\end{tcolorbox}

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Craig: GMU and many other universities allow you to upload your published papers and other writing samples. In many cases, even if the papers were not published at top places, we can still determine their quality by simply skimming over the paper.  
\end{tcolorbox}

@@

Finally, this is something very easy to do, but is missed by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply here? talk about a couple of professors who you're interested in working with (in many cases your application will be forwarded to them for evaluation).
Be careful not to send wrong statement to wrong school or mixing
facts (e.g., talking about school X but mentioned about working with
profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
should.

@@

Graduating from top universities \emph{that we recognize} helps.
However, if committee members do not know much about schools in your country, they will likely treat your school as
\emph{``unknown foreign''}, which can be a minus point (if your school is well-known, then it is \emph{``top foreign''}, which is definitely a plus).

So what can you do about this? several things such as asking your CS dept to put itself on CSRankings (it's the easiest way to get CS people to know about the school), explaining about your school in your statement (and asking your LOR writer to do that too), and of course, if you're Vietnamese, considering a CS PhD program that has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese professors}.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: Sometime PhD admission committee will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. I have looked at Vietnamese applications (whether they are assigned to me or not) and provide inputs to the reviewers of those applications, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{tcolorbox}

@@

This can be an issue for students in many top international universities where the competition is so high that very good students students can still have low ranking (and be overlooked by Admission committee).
So what to do with this? well, same as \hyperref[sec:your-school]{before}, e.g., put a note about this in your statement.

@@

You also do need to
take standardized English test. Just do well enough to pass minimum
requirement set by the university.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: The minimum for GMU CS (being above this might not mean much, but below is a \red{red flag}).
\begin{itemize}
\item GPA: $\ge 3.0$ in your undergrad (but as mentioned, we also consider the rank/prestige of your school)
\item GRE: not required (most CS graduate programs no longer require GRE, though it can help boost your profile)
%    \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
\item English requirement tests (one of the below)
  \begin{itemize}    
  \item TOEF: 88 pts in total AND $\ge 20$ points in each subsection OR
  \item IELTS: $\ge 6.5$ OR
  \item DuoLingo Graduate English: $\ge 120$ OR 
  \item Pearson Test of Academic English: $\ge 67$
  \end{itemize}  
\end{itemize}
\end{tcolorbox}

@@

An interview lasts about 15--30 mins, and one implicit thing you will be evaluated on is whether you can communicate effectively (i.e., speak/understand English).  You will also get chance to ask questions about the university so think of something to ask (just the same as you interview at a company).

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
At GMU, we are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some rare cases a faculty interviews a candidate that they see potentials and want argue for admission, i.e., without the interview, that candidate is definitely rejected. In any case, getting interview means you have a very good chance of being admitted.
\end{tcolorbox}

@@

Most schools will have an \textbf{Open House}, which is a great resource to learn about the school, department, faculty, research, living, etc. During the Open House, you get a chance to talk to individual faculty and current students.  Take notes of faculty who make you excited, count those that are taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisors, the dept, the area, funding situation etc.  Ask about anything you want to determine that they deserve \emph{you}.

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: GMU has Virtual Open House, e.g., \url{https://cs-gmu.github.io/cs-phd-voh-s23/}, which I've co-organized in the last two years. We invite all admitted PhD students to the VOH through Zoom to learn about the CS program, the department, GMU, and the DC area in general. Students also get opportunities to chat with professors and current students.
\end{tcolorbox}

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is the type of funding paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come the school, a company, or organization. Tab.~\ref{tab:funding} summarizes the differences.

\begin{table}
  \centering
  \caption{Different types of PhD funding}\label{tab:funding}
  \begin{tabular}{c|c|c|c}
    \toprule
    &\textbf{TA}&\textbf{RA}&\textbf{Fellowship}\\
    \midrule
    \textbf{From} & School & Profs. & School/External\\
    \textbf{For}                  & Teaching Assistant       & Research                        & Research                              \\
    \textbf{Tuition/Ins./Stipend} & Yes                      & Yes                             & Yes                                   \\
    \textbf{Cover Summer?}              & No                       & Maybe                           & Yes                                   \\
    \midrule
    \textbf{Pros}                 & Research Freedom         & Get to do research              & Research Freedom                      \\
    \textbf{Cons}                 & TA, Uncertain            & Research restriction, Uncertain & Competitive, limited             \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Graduating Assistantship (TA/RA)}
The most common type of funding is \textbf{graduate assistanship}, which is either TA or RA. Both TA and RA come with tuition waiving (you don't have to pay tuition), health insurance (this takes care of your insurance, which is a must have in the US), and most importantly, your stipend (i.e., your salary). Some universities also pay insurance for spouse/children (or give very good discount).

Several about stipend. First, the amount of stipend depends on the university, which in turns depend on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, a school year is (typically) 9-month in the US, so stipend is for 9 months (so divide by 9 for each month). Third, like for most source of income in the US, you will have to pay tax on your stipend. Finally, private universities might pay more for stipend.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
At GMU, TA and RA have similar benefits in tuition waiving and insurance.  For stipend, depending on the college and department, a 9-month graduate assistant stipend is set.  TA and RA will usually be that amount (TA will definitely be that, RA might fluctuate a bit depending on the stage of the student and the prof.). 
\end{tcolorbox}

@@

TA is common in the beginning when you haven't found your advisor who would pay you RA. As a TA, you help professors with their classes (e.g., grading or teaching labs/recitation). Your TAship is paid through the department, i.e., they hire you to help teach.  During a semester, a TA might work with several courses and professors (not necessary their advisor).  TA funding typically is not available during the summer, which has no school.

\paragraph{How to get TA?}  Unless you have other funding such as RA or Fellowships, TA is typically a default thing. When you apply to be a full-time student, you also sate that you need financial assistant. It is common that the PhD committee will either admit you and give you GTA, or reject you; i.e., we do not admit a student without supporting them.

@@

Even if you have other funding and do not need TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for research career where you often have to give talks and tell people about your work. Note that GMU sometimes has classes that a more senior student can teach.  In that case, you will be paid as a lecturer, which is higher than GTA.  This is a good opportunity for students to get teaching experience and also get paid more.
\subsection{Research Assistant (RA)}
RA is provided through a professor through their own funding so you can work on their project.  
You do not need to teach as an RA, so you can focus on your research. Depending on the professor, RA may be available during the summer.

\textbf{How to get RA?} When a professor recruits you, they will likely give you RA right away (e.g., when you apply).  A common scenario is that you first get admitted with TA, and then after a year or two find an advisor to support you with RA.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
Vu: If you're lucky and got recruited by a prof. who would give you RA right away, it's very likely you will get admitted.  For example, if a prof., even if not in PhD admission committee, wants to work with and funds you, the PhD admission committee will respect that decision and admit your application (unless your application has many red flags).
\end{tcolorbox}

@@

Fellowship is another type of funding in which the student applies for (e.g., from school, industries, government). Fellowships are typically competitive and generous, and gives pretty much all benefits tuition/insurance that a TA/RA has.  Moreover, it often gives higher stipend (including summer) and opens doors for job opportunities (e.g., internship).  For example, a student with a Microsoft fellowship will likely get an internship at Microsoft.

@@

\textbf{How to get Fellowship?} You apply for them.  The US government has many fellowships but these would likely require US citizenship or residency.  However, tech companies including Google, Microsoft, Facebook, IBM have fellowships that international students can apply for.

Prestigious fellowships typically require a clear and good research plan, so it is a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competiting with the top Ph.D. students at top universities worldwide.

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
At GMU, Ph.D. applicants are automatically eligible for a Presidential Fellowship.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you are not depending on any prof. or TA duties).  PhD admission committee members nominate applicants for this fellowship and the committee will vote and give the fellowship to the top 2.
\end{tcolorbox}

@@

\begin{enumerate}
    \item   Is an MS degree required for admission to PhD?
    \begin{description}
    \item No. In fact, student with BS can get MS degree "along the way" to PhD.  However, MS can help if it gives research experience or is from a more well-known school than your undergrad institution.
    \end{description}

    \item Can I apply to PhD to CS if my undergrad was in Math, Biology, Finance, etc?
    \begin{description}
        \item Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned.
    \end{description}

    \item What can you do to increase your admission chance?
    \begin{description}
        \item Show something that makes you \textbf{stand out}, e.g., are you a female or a minority in CS (research for this on Google)? Do you participate in outreach activities that help increase diversity and inclusion in CS?  All of these are important in CS in the US.

@@

    \item How do I address a professor Firstname Lastname?
    \begin{description}
        \item If you don't know that professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste the names, so no need to do so,  just Prof. or Dr. Lastname.

@@

\begin{tcolorbox}[left=1pt,right=1pt,top=1pt,bottom=1pt]
    Vu: I've been called Dr. Vu and I find it a bit amusing but am totally fine with it.
    \end{tcolorbox}
    \end{description}
\end{enumerate}

\section{Acknowledgement}

The following people contributed to this document: Craig Yu (GMU), Hakan Aydin (GMU).
